Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data

### 2019-05-15 thesis.tex
Get styling and bookmarks in line with department requirements

% Times New Roman for the thesis submission
% \usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{authblk}
\usepackage{url}
\usepackage{hyperref}
\usepackage[inline,shortlabels]{enumitem}
\usepackage[square,numbers]{natbib}
% \usepackage{natbib}
\usepackage[nomarkers,nolists,tabhead,fighead]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{changepage}
\bibliographystyle{unsrtnat}

@@

\pagenumbering{roman}
% Cover page
\begin{titlingpage}
	\begin{center}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\doublespacing \thetitle\\ \singlespacing
			\end{center}
		\end{adjustwidth}
		\vfill
		\theauthor\\
		\vfill
		A thesis submitted to the faculty of\\
		Brigham Young University\\
		in partial fulfillment of the requirements for the degree of\\
		\vspace{1em}
		Master of Science\\
		\vfill
		Stephen R. Piccolo, Chair\\
		Mark Clement\\
		David Wingate\\
		\vfill
		Department of Biology\\
		\vspace{1em}
		Brigham Young University
		\vfill
		Copyright \textcopyright \ \thedate \ \theauthor\\
		\vspace{1em}
		All Rights Reserved
	\end{center}

@@

% Title / abstract page
\begin{titlingpage}
	\begin{center}
		ABSTRACT\\
		\vspace{1em}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\thetitle\\
			\end{center}
		\end{adjustwidth}
		\vspace{1em}
		\theauthor\\
		Department of Biology, BYU\\
		Master of Science
	\end{center}
	% \begin{abstract}
		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
		However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
		For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
		known covariates that are not of interest to the study may have strong effects,
		and there may be large systemic effects when integrating multiple expression datasets.
		Additionally, many of these confounding effects represent higher-order interactions and cannot be removed with linear methodologies.
		We created Confounded in order to remove these effects from expression data.
		Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
		We tested the model on artificially constructed data and commonly used real gene expression datasets and compared the output with the output from other common batch adjustment algorithms.
		We also applied the model to remove the cancer-type-specific signal from pan-cancer expression data.
		Our software is available at \url{https://github.com/jdayton3/Confounded} to enable other scientists to remove confounding effects from their own data using our methodologies.
	% \end{abstract}
	\vfill
	\noindent
	Keywords: batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder
\end{titlingpage}

@@

\begin{titlingpage}
	\begin{center}
		ACKNOWLEDGMENTS\\
		This page is optional.
	\end{center}
\end{titlingpage}

\setcounter{page}{4}
\tableofcontents
\listoftables
\listoffigures
\pagebreak

@@

Gene expression data can be applied in many ways to advance our understanding of medicine and biology.
For example, expression data has been applied to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict the clinical outcome in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
Because of the sensitive nature of this data (i.e. gene product levels in organisms may respond drastically to small environmental changes) and of the measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
Three examples of this include
\begin{enumerate*}[(a)]
	\item batch effects, \label{it:batch}
	\item known covariates, and \label{it:cov}
	\item dataset-level effects. \label{it:int}
\end{enumerate*}
\ref{it:batch} Batch effects occur when expression data are generated in multiple runs or multiple batches, and slight systemic differences occur during the different runs, such as different technicians operating the machine or slight temperature differences in the room.
Batch effects are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
\ref{it:cov} Systemic bias can be even more pronounced when there are known covariates---for example, using data collected from different cancer types when cancer type is not of interest to the study \citep{dayton_classifying_2017-1}.
\ref{it:int} Finally, though batch effects within a dataset collected all by one lab at one time may be somewhat decreased experimentally by careful replication of experimental conditions, this is no longer possible when integrating different datasets;
however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see Figure \ref{fig:workflow}).

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
	\caption[Batch adjustment steps]{\textbf{Batch adjustment justification and steps}
	\begin{enumerate*}[(a)]
		\item When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
	\end{enumerate*}
	}
	\label{fig:workflow}
\end{figure}

Several methods exist for removing batch effects from gene expression datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear adjustments to remove batch effects, they cannot account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
However, these recent programs have several items that complicate their usability on many real-world datasets:
they require that the input data only contains two batches, that the batches are sufficiently large, and that the batches are perfectly balanced.
These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
	\caption[Network architecture of Confounded]{\textbf{Network architecture} of Confounded.
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

\begin{table}
	\centering
	\input{figures/final/datasets.tex}
	\caption[Dataset information]{\textbf{Dataset information} for each dataset used.}
	\label{tab:datasets}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
	\caption[Principal components analysis (PCA)]{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
	None of the datasets appear to be linearly separable.
	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
	\label{fig:tsne}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption[Mean squared error]{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
	MSE for unadjusted data will always be 0 because the input data is identical to the output data.
	Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mse}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption[Maximum mean discrepancy]{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
	In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mmd}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} for several datasets and adjusters.
	The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
	See also Figure \ref{fig:batch}.}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters.
	After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
	See also Figure \ref{fig:true_class}.}
	\label{tab:true_class}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mnist.png}
	\caption[MNIST handwritten digits]{\textbf{MNIST handwritten digits}
	\begin{enumerate*}[(a)]
		\item before any adjustment,
		\item with artificial noise added,
		\item adjusted for noise by the scale adjuster,
		\item adjusted for noise by ComBat, and
		\item adjusted for noise by Confounded.
	\end{enumerate*}
	Although Confounded seems to remove more noise from the background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}

@@

\bibliography{references}

>>>>

% Times New Roman for the thesis submission
% \usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{authblk}
\usepackage{url}
\usepackage[inline,shortlabels]{enumitem}
\usepackage[square,numbers]{natbib}
% \usepackage{natbib}
\usepackage{floatrow}
\floatsetup[table]{capposition=top}
\usepackage[nomarkers,nolists,tabhead,fighead,tablesfirst]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{changepage}
\usepackage{tocloft}
\usepackage[bookmarksnumbered,bookmarksopen=true,bookmarks=true]{hyperref}
\usepackage{bookmark}

% stuff for restyling the table of contents
\renewcommand\cftsecfont{\mdseries}
\renewcommand\cftsecpagefont{\mdseries}
\renewcommand{\contentsname}{\begin{center}\normalsize{\textnormal{TABLE OF CONTENTS}}\end{center}}
\renewcommand{\listtablename}{\begin{center}\normalsize{\textnormal{LIST OF TABLES}}\end{center}}
\renewcommand{\listfigurename}{\begin{center}\normalsize{\textnormal{LIST OF FIGURES}}\end{center}}

\bibliographystyle{unsrtnat}

@@

\pagenumbering{roman}
\begin{titlingpage}
	\setcounter{page}{1}
	\addcontentsline{toc}{section}{Title Page}
	\begin{center}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\doublespacing \thetitle\\ \singlespacing
			\end{center}
		\end{adjustwidth}
		\vfill
		\theauthor\\
		\vfill
		A thesis submitted to the faculty of\\
		Brigham Young University\\
		in partial fulfillment of the requirements for the degree of\\
		\vspace{1em}
		Master of Science\\
		\vfill
		Stephen R. Piccolo, Chair\\
		Mark Clement\\
		David Wingate\\
		\vfill
		Department of Biology\\
		\vspace{1em}
		Brigham Young University
		\vfill
		Copyright \textcopyright \ \thedate \ \theauthor\\
		\vspace{1em}
		All Rights Reserved
	\end{center}

@@

% Title / abstract page
\begin{titlingpage}
	\setcounter{page}{2}
	\phantomsection
	\addcontentsline{toc}{section}{Abstract}
	\begin{center}
		ABSTRACT\\
		\vspace{1em}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\thetitle\\
			\end{center}
		\end{adjustwidth}
		\vspace{1em}
		\theauthor\\
		Department of Biology, BYU\\
		Master of Science
	\end{center}
	% \begin{abstract}
		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
		However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
		For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
		known covariates that are not of interest to the study may have strong effects,
		and there may be large systemic effects when integrating multiple expression datasets.
		Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
		We created Confounded to remove these effects from expression data.
		Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
		We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
		We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
		Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
	% \end{abstract}
	\vfill
	\noindent
	Keywords: batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder
\end{titlingpage}

@@

\begin{titlingpage}
	\setcounter{page}{3}
	\phantomsection
	\addcontentsline{toc}{section}{Acknowledgments}
	\begin{center}
		ACKNOWLEDGMENTS\\
		TODO: Fill this out.
	\end{center}
\end{titlingpage}

\setcounter{page}{4}
\phantomsection
\addcontentsline{toc}{section}{Table of Contents}
\tableofcontents
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\newpage

% make bookmarks for Figure and Tables sections
\renewcommand{\figuresection}{\pdfbookmark[0]{Figures}{figure}Figures}
\renewcommand{\tablesection}{\pdfbookmark[0]{Tables}{table}Tables}

@@

Gene expression data can be applied in many ways to advance our understanding of medicine and biology.
For example, expression data has been applied to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict the clinical outcome in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
Because of the sensitive nature of this data (i.e. gene product levels in organisms may respond drastically to small environmental changes) and of the equipment used to quantify expression levels, expression data often contains unwanted confounding effects that can skew analysis.
Three examples of this include
\begin{enumerate*}[(a)]
	\item batch effects, \label{it:batch}
	\item known covariates, and \label{it:cov}
	\item dataset-level effects. \label{it:int}
\end{enumerate*}
\ref{it:batch} Batch effects occur when expression data are generated in multiple runs or multiple batches, and slight systemic differences occur during the different runs, such as different technicians operating the machine or slight temperature differences in the room.
Batch effects are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
\ref{it:cov} Systemic bias can be even more pronounced when there are known covariates---for example, using data collected from different cancer types when cancer type is not of interest to the study \citep{dayton_classifying_2017-1}.
\ref{it:int} Finally, though batch effects within a dataset collected all by one lab at one time may be somewhat decreased experimentally by careful replication of experimental conditions, this is no longer possible when integrating different datasets;
however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see Figure \ref{fig:workflow}).

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:workflow}}{workflow}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
	\begin{enumerate*}[(a)]
		\item When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
	\end{enumerate*}
	}
	\label{fig:workflow}
\end{figure}

Several methods exist for removing batch effects from gene expression datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear methods to model batch effects, they are not designed to account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
However, these recent programs have several items that complicate their usability on many real-world datasets:
they require that the input data only contains two batches, that the batches are sufficiently large, and that the batches are perfectly balanced.
These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.

@@

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:network}}{network}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
	\caption[Network architecture of Confounded]{\textbf{Network architecture} of Confounded.
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:training_loss}}{training_loss}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

\begin{table}
	\centering
	\input{figures/final/datasets.tex}
	\caption[Dataset information]{
		\pdfbookmark[1]{Table \ref{tab:datasets}}{datasets}
		\textbf{Dataset information} for each dataset used.
	}
	\label{tab:datasets}
\end{table}

@@

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:pca}}{pca}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
	\caption[Principal components analysis (PCA)]{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
	None of the datasets appear to be linearly separable.
	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:tsne}}{tsne}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding} (t-SNE) plot for the GSE37199 dataset before and after adjustment with several algorithms.
	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
	\label{fig:tsne}
\end{figure}

@@

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:mse}}{mse}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption[Mean squared error (MSE)]{\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
	Lower MSE represents that the adjuster has more faithfully reproduced the input data.
	MSE for unadjusted data will always be 0 because the input data is identical to the output data.
	Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mse}
\end{figure}
\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:mmd}}{mmd}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption[Maximum mean discrepancy (MMD)]{\textbf{Maximum mean discrepancy} (MMD) between different batches. Lower MMD indicates that the distributions of the different batches are more similar.
	In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
	In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.}
	\label{fig:mmd}
\end{figure}

@@

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:batch}}{batch}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption[Batch classification accuracy]{
		\pdfbookmark[1]{Table \ref{tab:batch}}{batch}
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also Figure \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:true_class}}{true_class}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{
		\pdfbookmark[1]{Table \ref{tab:true_class}}{true_class}
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also Figure \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:mnist}}{asdf}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mnist.png}
	\caption[MNIST handwritten digits]{\textbf{MNIST handwritten digits}
	\begin{enumerate*}[(a)]
		\item before any adjustment,
		\item with artificial noise added,
		\item adjusted for noise by the scale adjuster,
		\item adjusted for noise by ComBat, and
		\item adjusted for noise by Confounded.
	\end{enumerate*}
	Although Confounded seems to remove more noise from the background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}

@@

\newpage
\phantomsection
\addcontentsline{toc}{section}{References}
\bibliography{references}
